### 2017-08-18 reports/thesis/book.tex
Added compound Poisson normal approximation to thesis.

@@

\DeclareMathOperator{\expfamily}{ExpFamily}
\DeclareMathOperator{\expectation}{\mathbb{E}}
\DeclareMathOperator{\variance}{\mathbb{V}ar}
\DeclareMathOperator{\cov}{\mathbb{C}ov}
\DeclareMathOperator{\corr}{\mathbb{C}orr}
\DeclareMathOperator{\bernoulli}{Bernoulli}
\DeclareMathOperator{\betaDist}{Beta}
\DeclareMathOperator{\dirichlet}{Dir}
\DeclareMathOperator{\bin}{Bin}
\DeclareMathOperator{\MN}{Multinomial}
\DeclareMathOperator{\prob}{\mathbb{P}}
\DeclareMathOperator{\trace}{Tr}
\DeclareMathOperator{\normal}{N}
\DeclareMathOperator{\gammaDist}{Gamma}
\DeclareMathOperator{\poisson}{Poisson}
\DeclareMathOperator{\CPoisson}{CP\Gamma}

@@

Estimating $p$ is difficult and various methods were discussed by \cite{zhang2013likelihood}. One easy way is to to estimate $\mu$ and $\phi$ on a grid of $p$'s and then select the $p$ which maximises the likelihood \citep{dunn2005series}.

@@

\chapter{Compound Poisson}
\chapter{Inference}

>>>>

\setcounter{tocdepth}{2}
\setcounter{secnumdepth}{2}

@@

\DeclareMathOperator{\expfamily}{ExpFamily}
\DeclareMathOperator{\expectation}{\mathbb{E}}
\DeclareMathOperator{\variance}{\mathbb{V}ar}
\DeclareMathOperator{\cov}{\mathbb{C}ov}
\DeclareMathOperator{\corr}{\mathbb{C}orr}
\DeclareMathOperator{\bernoulli}{Bernoulli}
\DeclareMathOperator{\betaDist}{Beta}
\DeclareMathOperator{\dirichlet}{Dir}
\DeclareMathOperator{\bin}{Bin}
\DeclareMathOperator{\MN}{Multinomial}
\DeclareMathOperator{\prob}{\mathbb{P}}
\DeclareMathOperator{\trace}{Tr}
\DeclareMathOperator{\normal}{N}
\DeclareMathOperator{\gammaDist}{Gamma}
\DeclareMathOperator{\poisson}{Poisson}
\DeclareMathOperator{\CPoisson}{CP\Gamma}
\DeclareMathOperator{\normDist}{N}

@@

Estimating $p$ is difficult and various methods were discussed by \cite{zhang2013likelihood}. One easy way is to estimate $\mu$ and $\phi$ on a grid of $p$'s and then select the $p$ which maximises the likelihood \citep{dunn2005series}.

@@

\chapter{Compound Poisson}
In this chapter, the compound Poisson-Gamma distribution is reviewed and studied. Firstly methods for evaluating the density of the compound Poisson-Gamma distribution are discussed. Approximations such as the Normal approximation or saddle point approximation \citep{daniels1954saddlepoint} can be used. These approximate densities can be written in closed form. Exact methods can be computationally challenging when dealing with the infinite sum but can be done when summing over important terms \citep{dunn2005series}. Next, a method for parameter estimation will be proposed here using the EM algorithm \citep{dempster1977maximum}. However in the end of the chapter, it will be shown that no unique maximum likelihood estimator exist when certain parameters of the compound Poisson-Gamma distribution gets too large.

\section{Density Evaluation}
Suppose $X\sim\CPoisson(\lambda,\alpha,\beta)$ and $\lambda,\alpha,\beta>0$ then the p.d.f.~of $X$ is
\begin{equation}
p_X(x) = 
\delta(x) \euler^{-\lambda}
+
\euler^{-\beta x-\lambda}\frac{1}{x}\sum_{y=1}^{\infty}W_y
\quad\text{for }x\geqslant 0
\end{equation}
where
\begin{equation}
W_y = \frac{\beta^{y\alpha}\lambda^yx^{y\alpha}}{\Gamma(y\alpha)y!} \ .
\end{equation}
The evaluation of this p.d.f.~is difficult because the sum of $W_y$ cannot be simplified. However the m.g.f.~of $X$ can be written down in closed form
\begin{equation}
M(\theta)=\exp\left[\lambda\left(\left(\frac{\beta}{\beta-\theta}\right)^{\alpha}-1\right)\right] \ .
\end{equation}
The m.g.f.~allows the study of limiting distributions of the compound Poisson-Gamma distribution. These limiting distributions then can be used to approximate the p.d.f.~of a compound Poisson-Gamma random variable.

The saddle point approximation \citep{daniels1954saddlepoint} gives an approximate solution to inverting the Laplace transformation  of the m.g.f.~giving the p.d.f. However inverting the Fourier transformation of the characteristic function also gives the p.d.f.~using computational methods \citep{dunn2008evaluation}. The saddle point approximation will be studied here.

Lastly the infinite sum can be computationally summed in a clever way. This was done by summing only large terms in the sum and ignores small terms. By using Stirling's approximation, the largest term in the sum can be approximately found \citep{dunn2005series}.

\subsection{Normal Approxmation}
The m.g.f.~provides a starting point to what limiting distributions the compound Poisson-Gamma distribution converges to for large parameters. However the m.g.f.~of $X$ itself it not useful because when considering $\lambda\rightarrow\infty$ or $\alpha\rightarrow\infty$ then $M_X(\theta)\rightarrow\infty$. Also for $\beta\rightarrow\infty$, then $M_X(\theta)\rightarrow 1$ which is not useful either.

The compound Poisson-Gamma random variable $X$ should be standardised to obtain useful limiting results from the m.g.f. Let
\begin{equation}
Z = \frac{X-\expectation[X]}{\sqrt{\variance[X]}} \ ,
\end{equation}
then using previous results
\begin{equation*}
Z = \frac{\beta}{\sqrt{\alpha(\alpha+1)\lambda}}\left(
X-\frac{\alpha\lambda}{\beta}
\right)
\end{equation*}
\begin{equation}
Z = bX+a
\end{equation}
where
\begin{equation}
b = \frac{\beta}{\sqrt{\alpha(\alpha+1)\lambda}}
\end{equation}
and
\begin{equation}
a = -\sqrt{\frac{\alpha\lambda}{\alpha+1}} \ .
\end{equation}

The m.g.f.~of $Z$ is then
\begin{equation*}
M_Z(\theta)=\expectation\left[\euler^{Z\theta}\right]
\end{equation*}
\begin{equation*}
M_Z(\theta)=\expectation\left[\euler^{(bX+a)\theta}\right]
\end{equation*}
\begin{equation*}
M_Z(\theta)=\euler^{a\theta}M_X(b\theta) \ .
\end{equation*}
Substituting in the values
\begin{equation*}
M_Z(\theta)=\exp\left(-\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}\right)
\exp\left[
\lambda\left(
\left(
\frac{\beta}{\beta-\frac{\beta\theta}{\sqrt{\alpha(\alpha+1)\lambda}}}
\right)^\alpha-1
\right)
\right]
\end{equation*}
\begin{equation*}
M_Z(\theta)=\exp\left(-\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}\right)
\exp\left[
\lambda\left(
\left(
\frac{\sqrt{\alpha(\alpha+1)\lambda}}{\sqrt{\alpha(\alpha+1)\lambda}-\theta}
\right)^\alpha
-1\right)
\right]
\end{equation*}
or in a different form
\begin{equation}
M_Z(\theta)=\exp\left(-\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}\right)
\exp\left[
\lambda\left(
\left(
1-\frac{\theta}{\sqrt{\alpha(\alpha+1)\lambda}}
\right)^{-\alpha}
-1\right)
\right] \ .
\end{equation}

The above form is useful so that the binomial expansion can be conducted, that is
\begin{multline}
\left(
1-\frac{\theta}{\sqrt{\alpha(\alpha+1)\lambda}}
\right)^{-\alpha}
=
1+
\sum_{r=1}^\infty \frac{\theta^r\prod_{s=1}^r(\alpha+s-1)}{(\alpha(\alpha+1)\lambda)^{r/2}r!}
\\
\text{for }\frac{\|\theta\|}{\sqrt{\alpha(\alpha+1)\lambda}}<1 \ .
\end{multline}
Substituting in the binomial expansion
\begin{equation*}
M_Z(\theta)=\exp\left(-\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}\right)
\exp\left[
\lambda
\sum_{r=1}^\infty \frac{\theta^r\prod_{s=1}^r(\alpha+s-1)}{(\alpha(\alpha+1)\lambda)^{r/2}r!}
\right] \ .
\end{equation*}
Writing in full the $r=1,2$ terms
\begin{multline*}
M_Z(\theta)=\exp\left(-\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}\right)
\exp\left[
\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
+\frac{\theta^2}{2}
\right.\\\left.
+\lambda\sum_{r=3}^\infty \frac{\theta^r\prod_{s=1}^r(\alpha+s-1)}{(\alpha(\alpha+1)\lambda)^{r/2}r!}
\right]
\end{multline*}
and a term cancels out to get
\begin{equation}
M_Z(\theta)=
\exp\left[
	\frac{\theta^2}{2}
	+\sum_{r=3}^\infty
		\frac{\theta^r
			\prod_{s=1}^r(\alpha+s-1)
		}
		{(\alpha(\alpha+1))^{r/2}r!}
		\lambda^{1-r/2}
\right] \ .
\end{equation}

For large $\lambda$
\begin{equation}
\lim_{\lambda\rightarrow\infty}M_Z(\theta)=\exp\left[\frac{\theta^2}{2}\right]
\end{equation}
which is the same as the m.g.f.~of a standard Normal variable, therefore
\begin{equation}
\lim_{\lambda\rightarrow\infty}Z\sim\normal(0,1) \ .
\end{equation}
This should make sense as for high $\lambda$, the Poisson random variable has a high expectation, increasing the number of Gamma random variables in a summation. Increasing the number of terms in a summation will triggle the central limit theorem.

Next for high $\alpha$, it should be noted that
\begin{equation*}
\lim_{\alpha\rightarrow\infty}
	\frac{
		\prod_{s=1}^r(\alpha+s-1)
	}
	{(\alpha(\alpha+1))^{r/2}}
=
\frac{
		\prod_{s=1}^r\alpha
	}
	{\alpha^{r}}
\end{equation*}
so that
\begin{equation}
\lim_{\alpha\rightarrow\infty}
	\frac{
		\prod_{s=1}^r(\alpha+s-1)
	}
	{(\alpha(\alpha+1))^{r/2}} = 1 \ .
\end{equation}
As a result
\begin{equation}
\lim_{\alpha\rightarrow\infty}
M_Z(\theta)=
\exp\left[
	\frac{\theta^2}{2}
	+\sum_{r=3}^\infty
		\frac{\theta^r}
		{r!}
		\lambda^{1-r/2}
\right] \ .
\end{equation}
However this shows that taking the limit $\alpha\rightarrow\infty$ is not enough to get a Normal limiting distribution. For a Normal limiting distribution, the limit must be accompanied with the limit $\lambda\rightarrow\infty$, that is
\begin{equation}
\lim_{\lambda\rightarrow\infty}\lim_{\alpha\rightarrow\infty}Z\sim\normal(0,1) \ .
\end{equation}

Finally it should be noted that $M_Z(\theta)$ is independent of $\beta$. Thus $\beta$ will have no effect on the convergence to a Normal limiting distribution.

The limiting case where $\lambda\rightarrow 0$, $\alpha\rightarrow 0$ and $\beta\rightarrow 0$ will not be discussed here.

\subsection{Normal Approximation - General Case}
It was shown in certain situations, the compound Poisson-Gamma distribution can be approximated with the Normal distribution. This can be extended to the general compound Poisson.

Let $Y\sim\poisson(\lambda)$ where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be some random variable for i.i.d.~$i=1,2,3,\dotdotdot$ with m.g.f.~$M_U(\theta)$. Let
\begin{equation}
X|Y = \sum_{i=1}^{Y}U_i \ .
\end{equation}

It can be shown that the m.g.f.~of $X$ is
\begin{equation}
M_X(\theta)=\exp\left[
\lambda(M_U(\theta)-1)
\right] \ .
\end{equation}
As in the compound Poisson-Gamma case, it is useful to standardise $X$ using
\begin{equation*}
Z = \frac{X-\lambda\expectation[U]}{\sqrt{\lambda\expectation[U^2]}}
\end{equation*}
\begin{equation}
Z = \frac{X}{\sqrt{\lambda\expectation[U^2]}}
-\frac{\sqrt{\lambda}\expectation[U]}{\sqrt{\expectation[U^2]}} \ .
\end{equation}
The m.g.f.~of $Z$ is then
\begin{equation*}
M_Z(\theta)=\expectation\left[\euler^{Z\theta}\right]
\end{equation*}
\begin{equation*}
M_Z(\theta)=\exp\left[-\theta\frac{\sqrt{\lambda}\expectation[U]}{\sqrt{\expectation[U^2]}}\right]
\exp\left[
\lambda\left(
M_U\left(\frac{\theta}{\sqrt{\lambda\expectation[U^2]}}\right)-1
\right)
\right]
\end{equation*}

Taking the Taylor's series of $M_U(\theta)$
\begin{equation}
M_U(\theta) = 1 + \sum_{r=1}^\infty\frac{\theta^r}{r!}M_U^{(r)}(0)
\quad\text{for }\lim_{r\rightarrow\infty}
\left\|
\frac{\theta}{r+1}\frac{M_U^{(r+1)}(0)}{M_U^{(r)}(0)}
\right\|<1
\end{equation}
and substituting it into $M_Z(\theta)$
\begin{equation*}
M_Z(\theta)=\exp\left[-\theta\frac{\sqrt{\lambda}\expectation[U]}{\sqrt{\expectation[U^2]}}\right]
\exp\left[
\lambda\sum_{r=1}^\infty\frac{\theta^r}{(\lambda\expectation[U^2])^{r/2}r!}M_U^{(r)}(0)
\right] \ .
\end{equation*}
Writing out in full the $r=1,2$ terms
\begin{multline*}
M_Z(\theta)=\exp\left[-\theta\frac{\sqrt{\lambda}\expectation[U]}{\sqrt{\expectation[U^2]}}\right]
\exp\left[
\theta\frac{\sqrt{\lambda}M_U'(0)}{\sqrt{\expectation[U^2]}}
+
\frac{\theta^2M_U''(0)}{2\expectation[U^2]}
\right.\\+\left.
\lambda\sum_{r=3}^\infty\frac{\theta^r}{(\lambda\expectation[U^2])^{r/2}r!}M_U^{(r)}(0)
\right] \ .
\end{multline*}
By using the fact that $\expectation[U^r]=M_U^{(r)}(0)$ for $r=0,1,2\dotdotdot$, a few terms cancel out to get
\begin{equation}
M_Z(\theta)=\exp\left[
\frac{\theta^2}{2}
+
\sum_{r=3}^\infty\frac{\theta^r \expectation[U^r]}{(\expectation[U^2])^{r/2}r!}\lambda^{1-r/2}
\right] \ .
\end{equation}

For the limit $\lambda\rightarrow\infty$
\begin{equation}
\lim_{\lambda\rightarrow\infty}M_Z(\theta) = \exp\left[\frac{\theta^2}{2}\right]
\end{equation}
then it can be concluded that
\begin{equation}
\lim_{\lambda\rightarrow\infty}Z\sim\normal(0,1) \ .
\end{equation}
The convergence can be aided if the ratio $\dfrac{\expectation[U^r]}{(\expectation[U^2])^{r/2}}$ for $r=3,4,5,\dotdotdot$ are small.

\chapter{Inference}
